Fractals are Cool

### 2020-12-07 paper/paper_papaja.rmd
knitted .pdf
updates

author: 
  - name          : "Ben Chaloupka"
    affiliation   : "1"
    email         : "my@email.com"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
      - Grand Master Coding Wizard
  - name          : "Tesufuai Sameshima"
    affiliation   : "1"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
       - Writing - Original Draft Preparation
       - Writing - Review & Editing
       - Token Black Man
  - name          : "Scott Wallner"
    affiliation   : "1"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
       - Conceptualization
       - Writing - Original Draft Preparation
       - Writing - Review & Editing
       - Resident Fractal Expert

@@

authornote: |
 Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

@@

```{r setup, include = FALSE}
library(janitor)
library("papaja")
r_refs("r-references.bib")
# Sys.setenv(LANG = "en")
```

@@

# Introduction
  Fractal is a term originally derived to describe the “roughness,” (Mandelbrot) of objects, surfaces, or textures.  Fractal properties are pervasive throughout nature (Losa) and can be identified in the texture of clouds and mountains, in the branching pattern of trees and lightning, in the meandering shape of rivers and blood vessels.  These fractal properties are identified by their self-similarity and scale invariance. Self-similarity implies that a part of the structure maintains the same structure or pattern as the whole.  For example, if one were to zoom in on a small stretch of coastline, the jagged edges would look similar to the jaggedness of the coastline from a broader view.  Likewise, a small section of a fern is nearly identical to the branching shape of the entire fern.

@@

  To date there have been numerous research studies which explore the aesthetic preferences for fractals of different dimensions. Studies have included methods which present participants with a two alternative forced choice task of two fractals of differing dimension. Participants select the fractal image they prefer aesthetically and after viewing many pairs with differing dimensions, statistics are used to calculate the most preferred dimensionality range across participants. The overarching results of previous research have shown that humans, both adults and children prefer fractals within a range of dimensions that coincide with the fractal dimension that is found in natural objects and scenes, usually between 1.3 and 1.5.Remembering that fractal dimension is an analog for complexity, we can conclude that most humans prefer a moderate level of complexity compared to low or high complexity near the extremes. A secondary analysis of fractal fluency has shown that humans often also prefer fractal images to non-fractal images and that we prefer fractals which more closely resemble natural images to fractals which are geometrically exact. (Spehar, and Spies) Extensions on these research findings have led to the use of fractal images in therapeutic applications.  Studies have shown some benefits of experiencing fractals, including physiological responses in EEG measurement of brain activity while viewing fractals (Hagerhall), the reduction of stress using fractal art and architecture (Taylor) and even in the experience of hearing fractal melodies (Beauvois). There appears to be a clear link between, the fractal’s representation of complexity, the connection with natural objects and processes, and the human mind.

@@

## Participants
The study recruited 16 healthy young adults aging from 16-35 through the UO sona system.

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

@@

  In this study a dwell time paradigm which measures looking times of participants as they advance through a self paced slideshow, was used to measure the preferences for fractal images.  Slideshows were composed of a variety of different fractal types, each of which was presented in a sequential pattern of growth and decay of complexity.  Fractal images are abundant in nature, and fractal fluency theory (Taylor) has shown that humans have evolved perceptual preferences for fractal images. Previous studies have shown that both adults and children display patterns of aesthetic preference for a particular range of fractal dimension, a mathematical quantification analogous to complexity.  Work in these studies have found that preferences generally fall in a range of complexity which mimics the fractal dimension found in natural scenes.  Yet up to this point in time, these studies have utilized static and independent fractal images.  These studies also have generally employed an alternative forced choice paradigm to quantify participant’s preferences.  This study is novel in two ways.  First, we utilize the dwell time paradigm, a robust and reliable task which has been used in statistical learning and event processing studies in developmental psychology. (Baldwin, Kosie)  By measuring the time participants look at, or dwell on each image, patterns in preference, curiosity, or surprise can be mapped within or across individuals as they rapidly view many images sequentially.  We suggest that a dwell time paradigm provides a more nuanced insight into aesthetic preference or interest in a given stimuli set.  Results from this study have verified that this is a reliable measure to capture looking patterns and preference for fractal images.  The second way in which this study provides new findings to the fractal fluency literature is by utilizing dynamic fractals.  As discussed earlier, fractals pervade nature, and human perception has been tuned to process and prefer images that reflect fractals in nature.  But often our experiences of natural scenes are not static or independent, like the stimuli which has been used in all previous studies.  To better replicate natural experiences this study utilized a slideshow of fractal images which iterate through phases of growing and decaying complexity.  We sought to investigate whether participants would prefer dynamic fractals as they grow or decay in complexity.  The results show that there is a significant difference in the preference for dynamic fractals in the growth and decay phases, and that participants prefer to view fractals as they grow in complexity.  This can be seen in the increasing dwell times during the growth phase, and decreasing dwell times during the decay phase.  What is striking is that this pattern was consistent across 14 different fractal types and across a variety of number of iterations.

  There are a few limitations to the present study which inspire the future directions of this line of research.  Only 16 subjects, all college aged students, participated in the study.  In the future we plan to acquire more participants and include a larger range of ages.  After analyzing the data, a typical trend in dwell time paradigms emerged.  This pattern is that dwell times generally decrease as the slideshow proceeds.  As one would imagine this could occur because participants lose interest or simply increase the rate at which they proceed.  In order to alleviate this trend, we plan to use a fewer total number of fractal types and include attention checks to break up the presentation of the slide show. Data from this study which indicate the most preferential fractal types can be used to select the highest quality fractal sequences for subsequent studies. This should induce a more consistent display of attention across the entire task.  In this initial study all subjects were presented with fractal sequences which began with a growth phase followed by a decay phase.  We are interested to include a second condition in which a proportion of the participants view fractal sequences which start with a decay phase followed by the growth phase.  In future analysis of this study, and subsequent studies we also plan to analyze the same fractal images presented in a randomized order.  An important goal of this research trajectory is to more closely replicate the experiences of fractals in nature.  For this reason we can design studies which use 3 dimensional fractal images and another biologically inspired stimuli type cellular automata.  Surveys including the Autism Spectrum Quotient, and perceptual tasks like visual search and card sorting can be correlated with dwell time data can be utilized to better understand individual preferences and perceptual proclivities for complexity.  
  This study was the first of its kind, utilizing dynamic fractals presented in a dwell time paradigm.  The results from this study have shown that these methods are reliable and informative in regards to research questions which explore and connect fractal fluency, human perception, preference for complexity.

@@

>>>>

author: 
  - name          : "Ben Chaloupka"
    affiliation   : "1"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
      - Grand Master Coding Wizard
  - name          : "Tesufuai Sameshima"
    affiliation   : "1"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
       - Writing - Original Draft Preparation
       - Writing - Review & Editing
       - Token Black Man
  - name          : "Scott Wallner"
    affiliation   : "1"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
       - Conceptualization
       - Writing - Original Draft Preparation
       - Writing - Review & Editing
       - Resident Fractal Expert

@@

@@

```{r setup, include = FALSE}

if(!"tinytex" %in% rownames(installed.packages())) install.packages("tinytex")
tinytex::install_tinytex()
if(!"devtools" %in% rownames(installed.packages())) install.packages("devtools")
devtools::install_github("crsh/papaja")

library(janitor)
library("papaja")
r_refs("r-references.bib")
# Sys.setenv(LANG = "en")
```

@@

# Introduction
  Fractal is a term originally derived to describe the “roughness,” (@mandelbrot1989fractal) of objects, surfaces, or textures.  Fractal properties are pervasive throughout nature (@losa2009fractal) and can be identified in the texture of clouds and mountains, in the branching pattern of trees and lightning, in the meandering shape of rivers and blood vessels.  These fractal properties are identified by their self-similarity and scale invariance. Self-similarity implies that a part of the structure maintains the same structure or pattern as the whole.  For example, if one were to zoom in on a small stretch of coastline, the jagged edges would look similar to the jaggedness of the coastline from a broader view.  Likewise, a small section of a fern is nearly identical to the branching shape of the entire fern.

@@

  To date there have been numerous research studies which explore the aesthetic preferences for fractals of different dimensions. Studies have included methods which present participants with a two alternative forced choice task of two fractals of differing dimension. Participants select the fractal image they prefer aesthetically and after viewing many pairs with differing dimensions, statistics are used to calculate the most preferred dimensionality range across participants. The overarching results of previous research have shown that humans, both adults and children prefer fractals within a range of dimensions that coincide with the fractal dimension that is found in natural objects and scenes, usually between 1.3 and 1.5.Remembering that fractal dimension is an analog for complexity, we can conclude that most humans prefer a moderate level of complexity compared to low or high complexity near the extremes. A secondary analysis of fractal fluency has shown that humans often also prefer fractal images to non-fractal images and that we prefer fractals which more closely resemble natural images to fractals which are geometrically exact. (@spehar2003universal) Extensions on these research findings have led to the use of fractal images in therapeutic applications.  Studies have shown some benefits of experiencing fractals, including physiological responses in EEG measurement of brain activity while viewing fractals (@hagerhall2015human), the reduction of stress using fractal art and architecture (@taylor2006reduction) and even in the experience of hearing fractal melodies (@beauvois2007quantifying). There appears to be a clear link between, the fractal’s representation of complexity, the connection with natural objects and processes, and the human mind.

@@

##Participants
The study recruited 16 healthy young adults ages 16-35 through the UO SONA recruitment system.  All participants had normal or corrected to normal vision with no history of neurological damage.

##Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

@@

 A two-way ANOVA was conducted to examin the effect of dwell times on fractal presentation directions. The results indicated a significant interaction between the effects of fractal direction and dwell time, *F*(2,31.01) = 42.68, *p*<0.05). Simple main effects showed that growth conditions showed significantly longer dwell times compared to the decay or random directional conditions. 
```{r, fig.show = "hold", out.height =  "25%",fig.align = 'left'}

knitr::include_graphics(here::here("presentation_images","group1.png"))

  The following graph plots the dwell times over the course of the study for participant group 1. Group 1 and 2 where counterbalanced in fractal presentation order but show similar trends of decreasing dwell times as image number increases or the task itself progresses. Decreases in dwell times over time are likely attributed to fatigue or waning attention as the participants lose focus over time.      
```{r, fig.show = "hold", out.height =  "25%",fig.align = 'left'}

knitr::include_graphics(here::here("presentation_images","ovr_dwell_time2.png"))

@@

  In this study a dwell time paradigm which measures looking times of participants as they advance through a self paced slideshow, was used to measure the preferences for fractal images.  Slideshows were composed of a variety of different fractal types, each of which was presented in a sequential pattern of growth and decay of complexity.  Fractal images are abundant in nature, and fractal fluency theory (@taylor2006reduction) has shown that humans have evolved perceptual preferences for fractal images. Previous studies have shown that both adults and children display patterns of aesthetic preference for a particular range of fractal dimension, a mathematical quantification analogous to complexity.  Work in these studies have found that preferences generally fall in a range of complexity which mimics the fractal dimension found in natural scenes.  Yet up to this point in time, these studies have utilized static and independent fractal images.  These studies also have generally employed an alternative forced choice paradigm to quantify participant’s preferences.  This study is novel in two ways.  First, we utilize the dwell time paradigm, a robust and reliable task which has been used in statistical learning and event processing studies in developmental psychology. (@kosie2019attentiona)  By measuring the time participants look at, or dwell on each image, patterns in preference, curiosity, or surprise can be mapped within or across individuals as they rapidly view many images sequentially.  We suggest that a dwell time paradigm provides a more nuanced insight into aesthetic preference or interest in a given stimuli set.  Results from this study have verified that this is a reliable measure to capture looking patterns and preference for fractal images.  The second way in which this study provides new findings to the fractal fluency literature is by utilizing dynamic fractals.  As discussed earlier, fractals pervade nature, and human perception has been tuned to process and prefer images that reflect fractals in nature.  But often our experiences of natural scenes are not static or independent, like the stimuli which has been used in all previous studies.  To better replicate natural experiences this study utilized a slideshow of fractal images which iterate through phases of growing and decaying complexity.  We sought to investigate whether participants would prefer dynamic fractals as they grow or decay in complexity.  The results show that there is a significant difference in the preference for dynamic fractals in the growth and decay phases, and that participants prefer to view fractals as they grow in complexity.  This can be seen in the increasing dwell times during the growth phase, and decreasing dwell times during the decay phase.  What is striking is that this pattern was consistent across 14 different fractal types and across a variety of number of iterations.

  There are a few limitations to the present study which inspire the future directions of this line of research.  Only 16 subjects, all college aged students, participated in the study.  In the future we plan to acquire more participants and include a larger range of ages.  After analyzing the data, a typical trend in dwell time paradigms emerged.  This pattern is that dwell times generally decrease as the slideshow proceeds.  As one would imagine this could occur because participants lose interest or simply increase the rate at which they proceed.  In order to alleviate this trend, we plan to use a fewer total number of fractal types and include attention checks to break up the presentation of the slide show. Data from this study which indicate the most preferential fractal types can be used to select the highest quality fractal sequences for subsequent studies. This should induce a more consistent display of attention across the entire task.  In this initial study all subjects were presented with fractal sequences which began with a growth phase followed by a decay phase.  We are interested to include a second condition in which a proportion of the participants view fractal sequences which start with a decay phase followed by the growth phase.  In future analysis of this study, and subsequent studies we also plan to analyze the same fractal images presented in a randomized order.  An important goal of this research trajectory is to more closely replicate the experiences of fractals in nature.  For this reason we can design studies which use 3 dimensional fractal images and another biologically inspired stimuli type cellular automata.  Surveys including the Autism Spectrum Quotient, and perceptual tasks like visual search and card sorting can be correlated with dwell time data can be utilized to better understand individual preferences and perceptual proclivities for complexity.

  This study was the first of its kind, utilizing dynamic fractals presented in a dwell time paradigm.  The results from this study have shown that these methods are reliable and informative in regards to research questions which explore and connect fractal fluency, human perception, preference for complexity.
